A Logical Solution of the Linearity Challenge

### 2025-06-16 paper/calculus.tex
semantics

\begin{table}
    \caption{Operational semantics of \Calculus.}
    \[
        \begin{array}{rrcll}
        \SLink &
        \Link\x\y & \pcong & \Link\y\x \\
        \SComm &
        \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P} \\
        \SWait &
        \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} & x \ne y \\
        \SFail &
        \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
        \SSelect &
        \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} & x \ne y,z \\
        \SCase &
        \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} & x \ne y,z \\
        \SForkL &
        \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}\setminus\set{y,u,v} \\
        \SForkR &
        \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q}\setminus\set{y,u,v} \\
        \SJoin &
        \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
        \\
        \RLink &
        \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
        \RFail &
        \Cut\x{\Fail\y}{P} & \red & \Fail\y & x \ne y \\
        \RClose &
        \Cut\x{\Close\x}{\Wait\x.P} & \red & P \\
        \RSelect &
        \Cut\x{\Select\x{\Tag_i}\y.P}{\Case\x\y{Q_1}{Q_2}} & \red & \Cut\y{P}{Q_i} & i\in\set{1,2} \\
        \RFork &
        \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
        \RCut &
        \Cut\x{P}{R} & \red & \Cut\x{Q}{R} & P \red Q \\
        \RCong &
        P & \red & Q & P \pcong R \red Q \\
        \end{array}
    \]
\end{table}

>>>>

free and bound names

we consideer processes equal up to renaming of bound names

\begin{table}
    \caption{Operational semantics of \Calculus.}
    \label{tab:semantics}
    \[
        \begin{array}{rrcll}
        \SLink &
        \Link\x\y & \pcong & \Link\y\x \\
        \SComm &
        \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P} \\
        \SFail &
        \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
        \SWait &
        \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} & x \ne y \\
        \SSelect &
        \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} & x \ne y,z \\
        \SCase &
        \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} & x \ne y,z \\
        \SForkL &
        \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}\setminus\set{y,u,v} \\
        \SForkR &
        \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q}\setminus\set{y,u,v} \\
        \SJoin &
        \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
        \\
        \RLink &
        \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
        % \RFail &
        % \Cut\x{\Fail\y}{P} & \red & \Fail\y & x \ne y \\
        \RClose &
        \Cut\x{\Close\x}{\Wait\x.P} & \red & P \\
        \RSelect &
        \Cut\x{\Select\x{\Tag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
        \RFork &
        \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
        \RCut &
        \Cut\x{P}{R} & \red & \Cut\x{Q}{R} & P \red Q \\
        \RCong &
        P & \red & Q & P \pcong R \red Q \\
        \end{array}
    \]
\end{table}

The operational semantics of \Calculus is shown in \cref{tab:semantics} and is
given by two relations: a \emph{structural pre-congruence} relation $\pcong$ and
a \emph{reduction} relation $\red$. Structural pre-congruence relates processes
that we consider to be essentially indistinguishable. Reduction relates
processes in which a communication occurs. In some presentations of linear
logic~\cite{Doumane}, these relations are respectively dubbed \emph{external}
and \emph{internal} reductions. Let us describe the two relations more in
detail.

Structural pre-congruence is the least pre-congruence defined by rule
\rulename{s-*} rules.
%
\SLink and \SComm assert that links and parallel compositions are symmetric. The
remaining rules, when read from left to right, push a cut on $x$ underneath the
topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These rules
are key to float input/output actions to the top-level of a process, so that
they can interact with corresponding complementary actions in the context
surrounding the process.
%
Note that some rules have additional side-conditions preventing bound names to
become free.\Luca{Non proprio, sono tutti nomi bound.} Also, there are two
versions of \SForkL and \SForkR depending on which of the two continuations $P$
and $Q$ contains occurrences of the restricted channel $x$.
%
Notably and somewhat surprisingly, there is no pre-congruence rule asserting
that parallel composition is associative. While this is a sound and expected
property of parallel composition, it turns out to be inessential for proving all
of the properties of \Calculus that we consider. We leave it out in light of the
forthcoming formalization of the calculus, in which every simplication is
welcome.

The reduction rules \RLink, \RClose, \RSelect and \RFork are in one-to-one
correspondence with the well-known reductions of session calculis and the cut
reductions of linear logic. \RLink eliminates a link $\Link\x\y$ by unifying $x$
and $y$. \RClose closes a session $x$ when a termination signal is exchanged on
$x$. \RSelect and \RFork respectively model the communication of a tag $\Tag_i$
and of a fresh channel $y$. Since channels are linear in \Calculus, each
non-terminating communication carries along a \emph{fresh continuation} $z$ on
which the rest of the conversation may take place, in line with the semantics of
the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}.
%
The rule~\RCut propagates reductions through cuts and the rules \RCong closes
reduction by structural pre-congruence.
